---
title: "MATH446 Final Project"
author: "Kristin Fesmire, Connor Mennenoh, Veera Manohar Reddy Alavalapati"
date: "2023-05-03"
output: pdf_document
---

```{r}

rm(list = ls())
df <- read.csv('C:/Users/krtfe/Downloads/COVID-19_Daily_Cases__Deaths__and_Hospitalizations.csv')

# update the original df to reformatted date format
df$Date <- as.Date(df$Date, format = '%m/%d/%Y')

# order data by timew
dfT <- df[order(df$Date),]
dfRMNA <- na.omit(dfT[1:4])

# time series object
dfTime <- ts(dfRMNA, start = 1)

# segmenting the data
dfRMNA[dfRMNA$Date < '2020-03-26', 'Range'] <- 1
dfRMNA['2020-03-26' < dfRMNA$Date & dfRMNA$Date < '2021-08-20', 'Range'] <- 2
dfRMNA['2021-08-20' < dfRMNA$Date & dfRMNA$Date < '2022-02-28', 'Range'] <- 3
dfRMNA['2022-02-28' < dfRMNA$Date, 'Range'] <- 4
head(dfRMNA)

```



#### PACF and ACF for the three initial time series. 
#### ADF tests for the three initial time series also computed
\
```{r, warning = FALSE}

library(tidyr)
library(forecast)
library(tseries)
library(ggplot2)

# ACF and PACF for cases
acf(na.omit(df$Cases...Total), type = 'covariance', 
    main = 'ACF for Cases Time Series')
pacf(na.omit(df$Cases...Total), main = 'PACF for Cases Time Series')

# ACF and PACF for deaths
acf(na.omit(df$Deaths...Total), type = 'covariance', 
    main = 'ACF for Deaths Time Series')
pacf(na.omit(df$Deaths...Total), main = 'PACF for Deaths Time Series')

# ACF and PACF for hospitalizations
acf(na.omit(df$Hospitalizations...Total), type = 'covariance', 
    main = 'ACF for Hospitalizations Time Series')
pacf(na.omit(df$Hospitalizations...Total), 
     main = 'PACF for Hospitalizations Time Series')

# ADF test for stationarity, hospitalizations
cat('ADF Test for Cases:\n')
adf.test(dfRMNA$Hospitalizations...Total)

# ADF test for stationarity, cases
cat('\n\n\nADF Test for Deaths:\n')
adf.test(dfRMNA$Cases...Total)

# ADF test for stationarity, deaths
cat('\n\n\nADF Test for Hospitalizations:\n')
adf.test(dfRMNA$Deaths...Total)

# ADF test for stationarity, cases
cat('\n\n\nADF Test for Deaths, Differenced by 1:\n')
adf.test(diff(dfRMNA$Deaths...Total, 1))

```



#### Initial Analysis:
\
```{r}

library(ggcorrplot)

corrP2 <- cor(df[c(2:4)], method = 'pearson', use = 'pairwise.complete.obs')
ggcorrplot(corrP2, lab = TRUE,
           colors = c('cornflowerblue', 'white', 'darkorchid3'),
           title = 'Correlation Plot Comparing Hospitalizations, Deaths, and Cases')

# visualization of 2x2 plots
par(mfrow=c(2, 2))
hist(dfRMNA$Cases...Total, breaks = 100,
     main = 'Histogram of Total Cases',
     xlab = 'Cases', col = 'cornflowerblue')
hist(dfRMNA$Deaths...Total, breaks = 75,
     main = 'Histogram of Total Deaths',
     xlab = 'Deaths', col = 'cornflowerblue')
hist(dfRMNA$Hospitalizations...Total, breaks = 100,
     main = 'Histogram of Total Hospitalizations',
     xlab = 'Hospitalizations', col = 'cornflowerblue')

# visualization of 2x2 plots
par(mfrow=c(2, 2))
plot(dfRMNA$Date, dfRMNA$Cases...Total, type = 'l',
     ylab = 'Total Cases', xlab = 'Date',
     main = 'Total Cases, Time Series')
plot(dfRMNA$Date, dfRMNA$Deaths...Total, type = 'l',
     ylab = 'Total Deaths', xlab = 'Date',
     main = 'Total Deaths, Time Series')
plot(dfRMNA$Date, dfRMNA$Hospitalizations...Total, type = 'l',
     ylab = 'Total Hospitalizations', xlab = 'Date',
     main = 'Total Hospitalizations, Time Series')

# visualization of 2x2 plots, for the four submodels
par(mfrow=c(2, 2))
plot(dfRMNA[dfRMNA$Range == 1,]$Date, 
     dfRMNA[dfRMNA$Range == 1,]$Cases...Total, type = 'l',
     ylab = 'Total Cases', xlab = 'Date',
     main = 'Total Cases Time Series, Segment 1')
plot(dfRMNA[dfRMNA$Range == 2,]$Date, 
     dfRMNA[dfRMNA$Range == 2,]$Cases...Total, type = 'l',
     ylab = 'Total Cases', xlab = 'Date',
     main = 'Total Cases Time Series, Segment 2')
plot(dfRMNA[dfRMNA$Range == 3,]$Date, 
     dfRMNA[dfRMNA$Range == 3,]$Cases...Total, type = 'l',
     ylab = 'Total Cases', xlab = 'Date',
     main = 'Total Cases Time Series, Segment 3')
plot(dfRMNA[dfRMNA$Range == 4,]$Date, 
     dfRMNA[dfRMNA$Range == 4,]$Cases...Total, type = 'l',
     ylab = 'Total Cases', xlab = 'Date',
     main = 'Total Cases Time Series, Segment 4')



# paired ACF plots for the initial three time series
par(mfrow=c(2, 2))
pacf(na.omit(dfRMNA$Cases...Total), main = 'PACF, Cases')
pacf(na.omit(dfRMNA$Deaths...Total), main = 'PACF, Deaths')
pacf(na.omit(dfRMNA$Hospitalizations...Total), main = 'PACF, Hospitalizations')

```



#### ACF and PACF by segments
\
```{r, warning = FALSE}

# ACF and PACF for Cases, time segment 1
acf(na.omit(dfRMNA[dfRMNA$Range == 1,]$Cases...Total), 
    type = 'covariance', main = 'ACF for Cases Time Series, Segment 1')
pacf(na.omit(dfRMNA[dfRMNA$Range == 1,]$Cases...Total), 
     main = 'PACF for Cases Time Series, Segment 1')

# ACF and PACF for Cases, time segment 2
acf(na.omit(dfRMNA[dfRMNA$Range == 2,]$Cases...Total), 
    type = 'covariance', main = 'ACF for Cases Time Series, Segment 2')
pacf(na.omit(dfRMNA[dfRMNA$Range == 2,]$Cases...Total), 
     main = 'PACF for Cases Time Series, Segment 2')

# ACF and PACF for Cases, time segment 3
acf(na.omit(dfRMNA[dfRMNA$Range == 3,]$Cases...Total), 
    type = 'covariance', 
    main = 'ACF for Cases Time Series, Segment 3')
pacf(na.omit(dfRMNA[dfRMNA$Range == 3,]$Cases...Total), 
    main = 'PACF for Cases Time Series, Segment 3')

# ACF and PACF for Cases, time segment 4
acf(na.omit(dfRMNA[dfRMNA$Range == 4,]$Cases...Total), 
    type = 'covariance', 
    main = 'ACF for Cases Time Series, Segment 4')
pacf(na.omit(dfRMNA[dfRMNA$Range == 4,]$Cases...Total), 
    main = 'PACF for Cases Time Series, Segment 4')


# combined visualizations
par(mfrow=c(2, 2))
pacf(na.omit(dfRMNA[dfRMNA$Range == 1,]$Cases...Total), 
     main = 'PACF for Cases Time Series, Segment 1')
pacf(na.omit(dfRMNA[dfRMNA$Range == 2,]$Cases...Total), 
     main = 'PACF for Cases Time Series, Segment 2')
pacf(na.omit(dfRMNA[dfRMNA$Range == 3,]$Cases...Total), 
     main = 'PACF for Cases Time Series, Segment 3')
pacf(na.omit(dfRMNA[dfRMNA$Range == 4,]$Cases...Total), 
     main = 'PACF for Cases Time Series, Segment 4')


```



#### Plotting the qqnorm plots.
\
```{r, warning = FALSE}

# paired qqnorm plots for the first three time series
par(mfrow=c(1, 3))
qqnorm(dfRMNA$Cases...Total, main = 'Normal Q-Q Plot, Cases')
qqline(dfRMNA$Cases...Total)
qqnorm(dfRMNA$Deaths...Total, main = 'Normal Q-Q Plot, Deaths')
qqline(dfRMNA$Deaths...Total)
qqnorm(dfRMNA$Hospitalizations...Total, main = 'Normal Q-Q Plot, Hospitalizations')
qqline(dfRMNA$Hospitalizations...Total)

# paired qqnorm plots for the segmented cases time series
par(mfrow=c(2, 2))
qqnorm(dfRMNA[dfRMNA$Range == 1,]$Cases...Total, 
       main = 'Normal Q-Q Plot, Cases Segment 1')
qqline(dfRMNA[dfRMNA$Range == 1,]$Cases...Total)
qqnorm(dfRMNA[dfRMNA$Range == 2,]$Cases...Total, 
       main = 'Normal Q-Q Plot, Cases Segment 2')
qqline(dfRMNA[dfRMNA$Range == 2,]$Cases...Total)
qqnorm(dfRMNA[dfRMNA$Range == 3,]$Cases...Total, 
       main = 'Normal Q-Q Plot, Cases Segment 3')
qqline(dfRMNA[dfRMNA$Range == 3,]$Cases...Total)
qqnorm(dfRMNA[dfRMNA$Range == 4,]$Cases...Total, 
       main = 'Normal Q-Q Plot, Cases Segment 4')
qqline(dfRMNA[dfRMNA$Range == 4,]$Cases...Total)

```



#### ADF tests for the segmented time series. 
\
```{r, warning = FALSE}


# ADF test for stationarity, hospitalizations
cat('ADF Test for Cases, Segment 1:\n')
adf.test(na.omit(dfRMNA[dfRMNA$Range == 1,]$Cases...Total))

# ADF test for stationarity, cases
cat('\n\n\nADF Test for Cases, Segment 2:\n')
adf.test(na.omit(dfRMNA[dfRMNA$Range == 2,]$Cases...Total))

# ADF test for stationarity, deaths
cat('\n\n\nADF Test for Cases, Segment 3:\n')
adf.test(na.omit(dfRMNA[dfRMNA$Range == 3,]$Cases...Total))

# ADF test for stationarity, deaths
cat('\n\n\nADF Test for Cases, Segment 4:\n')
adf.test(na.omit(dfRMNA[dfRMNA$Range == 4,]$Cases...Total))


```



#### Basic Visualizations of the three time series
\
```{r}

plot(dfT$Date, dfT$Hospitalizations...Total, type = 'l',
     ylab = 'Total Amount of People', xlab = 'Date', 
     ylim = c(0, 10000), col = 3, main = 'Time Series, Plotted Together')
lines(dfT$Date, dfT$Cases...Total, type = 'l', col = 4)
lines(dfT$Date, dfT$Deaths...Total, type = 'l', col = 2)
legend('topleft', c('Hospitalizations', 'Cases', 'Deaths'), lty = 1,
       col = c(3, 4, 2))



plot(dfT$Date, dfT$Hospitalizations...Total, type = 'l',
     ylab = 'Total Amount of People', xlab = 'Date', 
     ylim = c(0, 500), col = 3, 
     main = 'Time Series, Plotted Together and Zoomed in')
lines(dfT$Date, dfT$Cases...Total, type = 'l', col = 4)
lines(dfT$Date, dfT$Deaths...Total, type = 'l', col = 2)
legend('topleft', c('Hospitalizations', 'Cases', 'Deaths'), lty = 1,
       col = c(3, 4, 2))
```



#### Lag information for each of the plots
\

```{r, warning = FALSE}
library(ggplot2)

# Lags for cases
hospTime <- window(ts(dfT, start = 1)[,c(1,2)])

# zoom out as much as possible, it'll be difficult to see otherwise
gglagplot(hospTime)

# Lags for deaths
hospTime <- window(ts(dfT, start = 1)[,c(1,3)])

# zoom out as much as possible, it'll be difficult to see otherwise
gglagplot(hospTime)

# Lags for hospitalizations
hospTime <- window(ts(dfT, start = 1)[,c(1,4)])

# zoom out as much as possible, it'll be difficult to see otherwise
gglagplot(hospTime)


```



#### Visualization of the dataset, for the introduction
\
```{r}

library(RGraphics)
library(gridExtra)
library(grid)

p1 <- tableGrob(head(dfT[1:4]))
grid.draw(p1)

```



#### Models for the three time series and the segmented time series
\
```{r}
library(MTS)

# models for each of the subsetted time periods and dates for cases
cat(paste('Time series for Total Cases:\n\n'))
casesARIMA <- auto.arima(dfRMNA$Cases...Total)
summary(casesARIMA)
cat(paste('\n\n\nTime series for Total Cases, time period 1:\n\n'))
casesUpARIMA <- auto.arima(dfRMNA[dfRMNA$Range == 1,]$Cases...Total)
summary(casesUpARIMA)
cat(paste('\n\n\nTime series for Total Cases, time period 2:\n\n'))
casesUp1ARIMA <- auto.arima(dfRMNA[dfRMNA$Range == 2,]$Cases...Total)
summary(casesUp1ARIMA)
cat(paste('\n\n\nTime series for Total Cases, time period 3:\n\n'))
casesUp2ARIMA <- auto.arima(dfRMNA[dfRMNA$Range == 3,]$Cases...Total)
summary(casesUp2ARIMA)
cat(paste('\n\n\nTime series for Total Cases, time period 4:\n\n'))
casesUp3ARIMA <- auto.arima(dfRMNA[dfRMNA$Range == 4,]$Cases...Total)
summary(casesUp3ARIMA)

# models for each of the subsetted time periods and dates for deaths
cat(paste('\n\n\n\nTime series for Total Deaths:\n\n'))
deathsARIMA <- auto.arima(dfRMNA$Deaths...Total, stationary = F)
summary(deathsARIMA)

# models for each of the subsetted time periods and dates for hospitalizations
cat(paste('\n\n\n\nTime series for Total Hospitalizations:\n\n'))
hospARIMA <- auto.arima(dfRMNA$Hospitalizations...Total)
summary(hospARIMA)

# dataframe for the errors:



```



#### Checking the residuals of the models
\
```{r}
library(MTS)
library(MASS)

# ARIMA for entire cases time series
checkresiduals(casesARIMA)

# ARIMA for segmented cases time series, segment 1
checkresiduals(casesUpARIMA)

# ARIMA for segmented cases time series, segment 2
checkresiduals(casesUp1ARIMA)

# ARIMA for segmented cases time series, segment 3
checkresiduals(casesUp2ARIMA)

# ARIMA for segmented cases time series, segment 4
checkresiduals(casesUp3ARIMA)

# ARIMA for entire cases time series
checkresiduals(deathsARIMA)

# ARIMA for entire cases time series
checkresiduals(hospARIMA)




```



#### Forecasting of the four submodels
\
```{r, warning = FALSE}

# paired forecast plots for the segmented cases time series
par(mfrow=c(2, 2))
plot(forecast(casesUpARIMA, h = 20), 
     main = 'Forecast of Cases, Segment 1')

plot(forecast(casesUp1ARIMA, h = 100), 
     main = 'Forecast of Cases, Segment 2')

plot(forecast(casesUp2ARIMA, h = 100), 
     main = 'Forecast of Cases, Segment 3')

plot(forecast(casesUp3ARIMA, h = 100), 
     main = 'Forecast of Cases, Segment 4')

```


